---
title: "ComutR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ComutR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(ComutR)
```

# Basic Comut plots

```{r basic comut, fig.width=8, fig.height=8}
input_maf <- data.frame(
  Tumor_Sample_Barcode = c("1", "1", "1", "2", "3", "4", "4"),
  Hugo_Symbol = c("A", "B", "C", "C", "A", "A", "B"),
  Variant_Classification = c("Missense_Mutation", "Nonsense_Mutation", "In_Frame_Del", "In_Frame_Del", "Missense_Mutation", "Nonsense_Mutation", "Nonsense_Mutation")
)

comut(data = input_maf)
```

# Preparing data for comut plots

```{r}
dat <- tibble::tribble(
  ~ID, ~GeneA, ~GeneB, ~GeneC,
  1, 1, 1, 1,
  2, 0, 0, 0,
  3, 1, 0, 0,
  4, 1, 1, 0
)
```

The main input to comut is a dataframe with columns found in a typical maf file. Three columns are required by comut:

- Tumor_Sample_Barcode
- Hugo_Symbol
- Variant_Classification

```{r}
input_data <- tibble::tribble(
  ~Tumor_Sample_Barcode, ~Hugo_Symbol, ~Variant_Classification,
  "1", "A", "Missense_Mutation",
  "1", "B", "Nonsense_Mutation",
  "1", "C", "In_Frame_Del",
  "2", "C", "In_Frame_Del",
  "3", "A", "Missense_Mutation",
  "4", "A", "Nonsense_Mutation",
  "4", "B", "Nonsense_Mutation"
  )
```


```{r fig.width=8, fig.height=8}
comut(data = input_data)
```


For more information on the maf format and column names, visit [maf format](https://docs.gdc.cancer.gov/Data/File_Formats/MAF_Format/)

# Adding metadata

Metadata can be added through the metadata argument of comut. This expects and dataframe with Tumor_Sample_Barcode and then additional columns for the metadata.

Along with metadata, you must pass the col_maps argument, which contains a named list of color maps corresponding to the columns in the metadata.

Here we have gender and biopsy site information for each sample, and we define color maps for the variables accordingly:

```{r}
meta <- tibble::tribble(
  ~Tumor_Sample_Barcode, ~Gender, ~Biopsy,
  "1", "Male", "Primary",
  "2", "Female", "Metastasis",
  "3", "Male", "Metastasis",
  #"4", "Female", "Primary"
)

# Define color mapping
gender_colors <- c(
  "Male" = "lightblue",
  "Female" = "pink"
)
biopsy_type_colors <- c(
  "Primary" = "lightgreen",
  "Metastasis" = "maroon"
)
color_maps <- list(
  "Gender" = gender_colors,
  "Biopsy" = biopsy_type_colors
)
```

Note that patients without metadata will be colored grey.

```{r fig.width=8, fig.height=8}
comut(data = input_data, metadata = meta, col_maps = color_maps)
```

# Adding Barplots

Barplot data is formatted in a similar manner to metadata. A column for Tumor_Sample_Barcode is required, and then columns for the barplot values. If a stacked barplot is desired (for mutation signatures for example) then there should be multiple columns corresponding to the proportion of each value.

Barplots are defined within a named list of named lists. The top level name of each sublist is used as the y-axis label for the barplot. Each sublist should contain "data" containing the barplot data, "colors" containing a color map for the barplot, and "legend" a boolean for whether a legend should be created.

```{r}
signatures <- tibble::tribble(
  ~Tumor_Sample_Barcode, ~Aging, ~Other,
  "1", 0.5, 0.5,
  "2", 0.25, 0.75,
  "3", 0.75, 0.25,
  "4", 1, 0
)

total_snp_data <- tibble::tribble(
  ~Tumor_Sample_Barcode, ~n_snps,
  "1", 10,
  "2", 4,
  "3", 2,
  "4", 7
)

signature_colors <- c(
  "Aging" = "coral1",
  "Other" = "antiquewhite"
)

# Barplot specifications
# Note: the order of the list is reverse of the order in the plot!
barplot_data <- list(
  "Mutation Signatures" = list(
    "data" = signatures,
    "colors" = signature_colors,
    "legend" = TRUE
    ),
  "N SNPs" = list(
    "data" = total_snp_data,
    "colors" = c("n_snps" = "cadetblue3"),
    "legend" = FALSE
    )
)
```

```{r fig.width=10, fig.height=10}
comut(
  data = input_maf,
  barplot_data = barplot_data
)
```

Putting all of the components together, we get something like this:

```{r fig.width=10, fig.height=10}
comut(
  data = input_maf,
  metadata = meta,
  col_maps = color_maps,
  barplot_data = barplot_data
)
```

# Text Annotations

ComutR allows the user to add arbitrary text annotations to the cells of the comut plot. This can be specified as a column name that appears in the input data. Here we show adding allele fraction annotations to each mutation:

```{r fig.width=8, fig.height=8}
input_data <- tibble::tribble(
  ~Tumor_Sample_Barcode, ~Hugo_Symbol, ~Variant_Classification, ~Allele_Fraction,
  "1", "A", "Missense_Mutation", 0.5,
  "1", "B", "Nonsense_Mutation", 0.25,
  "1", "C", "In_Frame_Del", 0.75,
  "2", "C", "In_Frame_Del", 0.6,
  "3", "A", "Missense_Mutation", 0.5,
  "4", "A", "Nonsense_Mutation", 0.35,
  "4", "B", "Nonsense_Mutation", 0.8
  )

comut(
  data = input_data,
  text_annotation = "Allele_Fraction"
  )
```

# Limiting samples or features

If a subset of samples or features in the input data are desired, the user can specify the ids or features_of_interest arguments respectively.

```{r fig.width=8, fig.height=8}
input_data <- tibble::tribble(
  ~Tumor_Sample_Barcode, ~Hugo_Symbol, ~Variant_Classification,
  "1", "A", "Missense_Mutation",
  "1", "B", "Nonsense_Mutation",
  "1", "C", "In_Frame_Del",
  "2", "C", "In_Frame_Del",
  "3", "A", "Missense_Mutation",
  "4", "A", "Nonsense_Mutation",
  "4", "B", "Nonsense_Mutation"
  )

comut(
  data = input_data,
  ids = c("1", "3", "4"),
  features_of_interest = c("A", "B")
  )
```
